Inferring the distribution and rate of Earth-like exoplanets \\ from noisy individual-planet inferences

### 2014-04-30 document/ms.tex
reworking model generalities and removing redundant sections

\newcommand{\nyu}{2}
\newcommand{\mpia}{3}
\newcommand{\princeton}{4}
\newcommand{\berkeley}{5}
\author{%
    Daniel~Foreman-Mackey\altaffilmark{1,\nyu},
    David~W.~Hogg\altaffilmark{\nyu,\mpia},
    Timothy~D.~Morton\altaffilmark{\princeton},
    Erik~A.~Petigura\altaffilmark{\berkeley},
    \etal
}
\altaffiltext{1}         {To whom correspondence should be addressed:
                          \url{danfm@nyu.edu}}
\altaffiltext{\nyu}      {Center for Cosmology and Particle Physics,
                          Department of Physics, New York University,
                          4 Washington Place, New York, NY, 10003, USA}
\altaffiltext{\mpia}     {Max-Planck-Institut f\"ur Astronomie,
                          K\"onigstuhl 17, D-69117 Heidelberg, Germany}
\altaffiltext{\princeton}{Department of Astrophysics, Princeton University,
                          Princeton, NJ, 08540, USA} % CHECK THIS ZIP CODE
\altaffiltext{\berkeley} {Department of Astronomy,
                          University of California at Berkeley,
                          Berkeley, CA, XXXXX, USA} % CHECK THIS ZIP CODE

@@

In this \paper, we are asking the question: \emph{what constraints can we put
on the occurence rate of exoplanets given all the light curves measured by
\kepler?}
Mathematically, this involves evaluating the so-called \emph{marginalized
likelihood function}
\begin{eqnarray}\eqlabel{crazylike}
p(\{\data_k\}\,|\,\ratepars) &=&
    \int\left[\prod_{k=1}^K p(\data_k\,|\,\entry_k)\right ]
    \,p(\{\entry_k\}\,|\,\ratepars)
    \dd\{\entry_k\}
\end{eqnarray}
where $\{\data_k\}$ is the set of all light curves, \ratepars\ is a vector of
parameters describing the occurence rate function $\rate_\ratepars(\entry)$
and \entry\ is a vector of physical parameters describing a planetary system
(period, radius ratio, \etc).

@@

A catalog is a dimensionality reduction of the data that it is based on where
all the (relevant) information has been retained (we hope!).
In the context of \kepler, the catalog reduces the set of downloaded pixel
time series (approximately XXXX for the average \kepler\ target) to
probabilistic constraints on a handful of physical parameters---we'll call
them \entry---like the orbital period and radius.
If we take this set of parameters as \emph{sufficient statistics} of the data
then we can, in theory, compute \eq{crazylike}---up to an unimportant
constant---without ever looking at the raw data again.
This is true because the catalog provides constraints on the posterior
probability of the parameters $\{\entry_k\}$ under some ``interim prior''
\interim
\begin{eqnarray}\eqlabel{crazypost}
p(\{\entry_k\}\,|\,\{\data_k\},\,\interim) &=&
\frac{p(\{\entry_k\}\,|\,\interim)}{p(\{\data_k\}\,|\,\interim)}\,
\prod_{k=1}^K p(\data_k\,|\,\entry_k) \quad.
\end{eqnarray}
We can then multiply the integrand in \eq{crazylike} by
\begin{eqnarray}
\frac{p(\{\entry_k\}\,|\,\{\data_k\},\,\interim)}
     {p(\{\entry_k\}\,|\,\{\data_k\},\,\interim)}
\end{eqnarray}
and use \eq{crazypost} to find
\begin{eqnarray}
\frac{p(\{\data_k\}\,|\,\ratepars)}{p(\{\data_k\}\,|\,\interim)} &=&
    \int
    \frac{p(\{\entry_k\}\,|\,\ratepars)}{p(\{\entry_k\}\,|\,\interim)}\,
    p(\{\entry_k\}\,|\,\{\data_k\},\,\interim)
    \dd\{\entry_k\} \quad.
\end{eqnarray}
On the right hand side of this equation, the data come in only through the
quantity $p(\{\entry_k\}\,|\,\{\data_k\},\,\interim)$ that is provided by the
catalog!

@@

\emph{Note: I think that we should use $T$ instead of $P$ for period because
$P$ is too heavily overloaded by the probabilities\ldots}

From our perspective, a ``model'' is a likelihood function---a probability
density (or distribution) function (PDF) for the data given
parameters---and a set of prior PDFs for the parameters.
In what follows, we will treat as ``data'' a catalog of exoplanet period
and radius measurements (\citealt{petigura}).
\emph{(I don't think need this next sentence).}
We will treat the uncertainties on those measurements as ``prior
information''; that is, we will treat those as true as given and not as being
generated by the parameterized likelihood function.

In the models that follow, the most important parameters will be parameters
of the PDF for the exoplanet periods and radii.
However, there will also be ``latent parameters'' that represent the ``true''
values of the exoplanet periods \period\ and radii \radius.
That is, for each exoplanet, there is a ``measured'' or ``observed'' period
\obs{\period}\ and radius \obs{\radius}, which constitute our data.
These relate to the unobserved ``true'' period and radius by a noise model.
For the conclusions of this study, the true values of period and radius are
nuisance parameters; we will marginalize them out.

Let's also introduce more succinct notation for a \emph{catalog}.
We'll refer to the set of period and radius measurements for a set of $K$
candidate exoplanets as
\begin{eqnarray}
\catalog &=& \{\entry_k\}_{k=1}^K
\end{eqnarray}
where each entry $\entry_k = (\period_k,\,\radius_k)$ is a tuple containing a
period and radius.

The purpose of this \paper\ is to determine constraints on the \emph{rate
function} of exoplanets
\begin{eqnarray}\eqlabel{true-rate}
\rate_\ratepars (\entry)
&=& \frac{\dd N}{\dd\ln\period\dd\ln\radius}
\end{eqnarray}
given our dataset.
Here we have not made any assumptions about the functional form of \rate, we
are only stating that it can be modeled by some---possibly very large---set of
parameters \ratepars.

In this \paper, we will assume that the completeness function of the survey is
known and it can be represented as a probability at every (true) period and
radius $\completeness (\entry)$.
This probability will also include the geometric transit probability as a
function of period $\transitprob (\period)$.
These selection effects can be incorporated into the rate from \eq{true-rate}
to get the \emph{observable rate}
\begin{eqnarray}\eqlabel{obs-rate}
\obs{\rate}_\ratepars (\entry)
&=& \rate_\ratepars(\entry)\,\completeness(\entry)\quad.
\end{eqnarray}

Given this parameterized rate $\obs{\rate}_\ratepars$, we can compute the
Poisson likelihood of a \catalog\ (\citealt{tabachnik,youdin,dong})
\begin{eqnarray}\eqlabel{true-like}
\ln p(\catalog\,|\,\ratepars) &\propto&
\sum_{k=1}^K \ln\obs{\rate}_\ratepars (\entry_k)
- \int \obs{\rate}_\ratepars (\entry) \dd\entry
\end{eqnarray}
where we have assumed that every object in \catalog\ is an independent
Poisson sample from the rate.

If we had perfect noiseless measurements of the periods and radii for every
object in our catalog, \eq{true-like} provides a metric for determining
constraints on the parameters of the rate function.
In practice, the measurements of \catalog\ are noisy so we need to include
these effects.

\section{Including observational uncertainties}

The likelihood function given in \eq{true-like} is a model of the
probability of observing a catalog with \emph{true} periods and radii.
If we want to include the effects of measurement uncertainties, we need to
take a moment to consider the meaning of measurement uncertainties.

The catalog of planet candidates considered in this \paper\
(\citealt{petigura}) was generated by drawing samples from a posterior
probability distribution for physical parameters (period, radius, impact
parameter, \etc) conditioned on \kepler\ light curves.
The $k$th entry in this catalog is, therefore, a representation of the
posterior PDF
\begin{eqnarray}\eqlabel{measurement}
p(\entry_k\,|\,\obs{\entry}_k,\,\interim)
\end{eqnarray}
using the notation from the previous section.
In \eq{measurement}, the notations \interim\ indicates that the measurement
was made with a specific choice of prior over parameters
$p(\entry_k\,|\,\interim)$.
We will refer to these priors as \emph{interim priors} because our goal is to
infer the distribution that is consistent with the data.

More generally, a published catalog represents a \emph{posterior PDF over
plausible catalogs}
\begin{eqnarray}\eqlabel{catalog-pdf}
p(\catalog\,|\,\obs{\catalog},\,\interim) \quad.
\end{eqnarray}
It is normally safe to assume that each entry in the catalog is conditionally
independent given the choices of interim priors.
\dfmtodo{Why and when is this a good/bad/necessary assumption?}
Under this assumption, \eq{catalog-pdf} factorizes as
\begin{eqnarray}\eqlabel{catalog-factor}
p(\catalog\,|\,\obs{\catalog},\,\interim) &=&
\prod_{k=1}^K p(\entry_k\,|\,\obs{\entry}_k,\,\interim) \quad.
\end{eqnarray}

It will be useful in the following discussion to represent the catalog PDF as
a set of $N$ samples
\begin{eqnarray}\eqlabel{catalog-samples}
\catalog^{(n)} &\sim& p(\catalog\,|\,\obs{\catalog},\,\interim) \quad.
\end{eqnarray}
For the general catalog, these could be generated using MCMC or a similar
sampling algorithm.
Under the conditional independence assumption of \eq{catalog-factor}, however,
the problem is much simpler.
Each entry $\entry_k$ can be sampled independently and then the catalog
samples $\catalog^{(n)}$ are given by the outer product of the independent
sample lists.

Now that we have a representation of the catalog, we want to infer constraints
on the occurrence rate of exoplanets conditioned on all the data.
To perform this inference, we need to compute the  \emph{marginalized
likelihood}
\begin{eqnarray}\eqlabel{marg-likelihood}
p(\obs{\catalog}\,|\,\ratepars) &=&
\int p(\catalog,\,\obs{\catalog}\,|\,\ratepars) \dd\catalog \quad.
\end{eqnarray}
The marginalized likelihood measures the probability of all the \kepler\ light
curves given a particular setting of the occurrence rate parameters.
If we can efficiently compute this then we can optimize to find the
maximum likelihood occurrence rate or we can apply priors $p(\ratepars)$ and
sample to find posterior constraints on the occurrence rate.
Note that any results obtained through \eq{marg-likelihood} will
\emph{correctly account for all the measurement uncertainties}.

The integral in \eq{marg-likelihood} cannot be computed analytically and, in
fact, it will generally be extremely computationally expensive to evaluate
numerically (\dfmtodo{CITE}).
In our case, however, we only need to evaluate this integral up to a constant
and there is a simple approximation---using the sampling representation of the
catalog---that can be sufficiently accurate for our purposes.
To derive this approximation, first recall that the joint distribution inside
the integral in \eq{marg-likelihood} factorizes as
\begin{eqnarray}
p(\catalog,\,\obs{\catalog}\,|\,\ratepars) &=&
p(\catalog\,|\,\ratepars)\,p(\obs{\catalog}\,|\,\catalog)
\end{eqnarray}
because the occurrence rate generates the \emph{true} distribution of periods
and radii.
Then, we can multiply the integrand by
\begin{eqnarray}
\frac{p(\catalog\,|\,\obs{\catalog},\,\interim)}
     {p(\catalog\,|\,\obs{\catalog},\,\interim)}
\end{eqnarray}
and simplify to find
\begin{eqnarray}
\frac{p(\obs{\catalog}\,|\,\ratepars)}
     {p(\obs{\catalog}\,|\,\interim)}
&=&
\int \frac{p(\catalog\,|\,\ratepars)}{p(\catalog\,|\,\interim)}\,
p(\catalog\,|\,\obs{\catalog},\,\interim)
\dd\catalog \quad.
\end{eqnarray}
Using the samples over catalogs $\catalog^{(n)}$ as defined above, we can
approximate this integral---up to an irrelevant normalization constant---as
\begin{eqnarray}\eqlabel{approx-marg-likelihood}
p(\obs{\catalog}\,|\,\ratepars) &\appropto&
\sum_{n=1}^N
\frac{p(\catalog^{(n)}\,|\,\ratepars)}{p(\catalog^{(n)}\,|\,\interim)}
\end{eqnarray}
where the numerator inside the sum is given by \eq{true-like} and the
denominator is the interim prior evaluated at the coordinates of the samples.

This is a generalization of the approximation presented previously in the
exoplanet literature (\citealt{hogge}) but in this case, the likelihood
function \eq{true-like} doesn't factorize over entries in the catalog.
As a result, the variance of \eq{approx-marg-likelihood} is more sensitive to
the choice of interim prior.
In practice, we have found that including the completeness function
$\completeness(\entry)$ in $p(\entry\,|\,\interim)$ greatly improves
performance.
This term can be incorporated into existing samplings---generated using less
informative priors---using rejection sampling.

\section{Parameterization of the occurrence rate}

In the following sections, we will model the occurrence rate
$\rate(\period,\,\radius)$ as a two dimensional grid with $N$ bins in
$\ln\period$ and $\ln\radius$.
We choose to fix the bin locations leaving only the bin heights
$\binheights = \{\binheight_i\}_{i=1}^{N}$ as free parameters.

\dfmtodo{What about the normalization?}

Since the choice of bin locations and number is arbitrary, the most
conservative strategy is to choose very small bins---yielding an effectively
non-parametric model---and choose a prior (or regularization) that encodes any
prior expectations about the distribution.
In particular, it seems reasonable to expect the distribution to be
``smooth''.
The most extreme version of this model would be a completely non-parametric
rate function---generated by, for example, a Gaussian process
(\citealt{poiss-gp})---but this discussion is beyond the scope of this
\paper.
Instead, we'll simply choose small bins and add a Gaussian process prior
(\dfmtodo{CITE}) on the logarithms of the bin heights
\begin{eqnarray}
\ln p(\binheights\,|\,\mu,\,\bvec{\smooth}) &=&
-\frac{1}{2}\,\bvec{r}^\mathrm{T}\,\bvec{K}^{-1}\,\bvec{r}
-\frac{1}{2}\ln \det \bvec{K} - \frac{N}{2}\ln 2\pi
\end{eqnarray}
where $\bvec{r}$ is the vector of residuals
\begin{eqnarray}
\bvec{r} &=& \left(\begin{array}{cccc}
\binheight_1-\mu & \binheight_2-\mu & \cdots & \binheight_N-\mu
\end{array}\right)^\mathrm{T}
\end{eqnarray}
and $\bvec{K}$ is the kernel matrix evaluated \emph{at the bin centers}
$\bincenter_i$
\begin{eqnarray}
K_{ij}(\bvec{\smooth}) &\equiv& \smooth_0^2\,\exp \left(
-\frac{1}{2}[\bincenter_i-\bincenter_j]^\mathrm{T}\,
\mathrm{diag}(1/\smooth_k^2)\,
[\bincenter_i-\bincenter_j] \right) \quad.
\end{eqnarray}

Below, we will explore the effects of this choice of prior but when we
generate our final results, we will actually apply hyperpriors to the
parameters $\mu$ and $\bvec{\smooth}$ and marginalize over their values.

@@

The largest injected radius is $16\,R_\oplus$ but, because of the measurement
uncertainties on the radii, it will be necessary for us to model the
distribution to larger radii.
To do this, we approximate the survey completeness for $\radius > 16\,R\oplus$
as 1.

\dfmtodo{Hopefully Erik will send us MCMC samples. If not, describe the
re-sampling procedure that we used. Either way, we'll need to re-sample under
the completeness probability for the purposes of the importance sampling.}

@@

\acknowledgments
It is a pleasure to thank
    \ldots
for helpful contributions to the ideas and code presented here.
This project was partially supported by the NSF (grant AST-0908357), and NASA
(grant NNX08AJ48G).
This research made use of the NASA \project{Astrophysics Data System}.

>>>>

\newcommand{\nyu}{2}
\newcommand{\mpia}{3}
\newcommand{\princeton}{4}
\newcommand{\berkeley}{5}
\author{%
    Daniel~Foreman-Mackey\altaffilmark{1,\nyu},
    David~W.~Hogg\altaffilmark{\nyu,\mpia},
    Timothy~D.~Morton\altaffilmark{\princeton},
    \etal
}
\altaffiltext{1}         {To whom correspondence should be addressed:
                          \url{danfm@nyu.edu}}
\altaffiltext{\nyu}      {Center for Cosmology and Particle Physics,
                          Department of Physics, New York University,
                          4 Washington Place, New York, NY, 10003, USA}
\altaffiltext{\mpia}     {Max-Planck-Institut f\"ur Astronomie,
                          K\"onigstuhl 17, D-69117 Heidelberg, Germany}
\altaffiltext{\princeton}{Department of Astrophysics, Princeton University,
                          Princeton, NJ, 08540, USA} % CHECK THIS ZIP CODE

@@

In this \paper, we are asking the question: \emph{what constraints can we put
on the occurence rate of exoplanets given all the light curves measured by
\kepler?}
Mathematically, this involves evaluating the so-called \emph{marginalized
likelihood function}
\begin{eqnarray}\eqlabel{crazylike}
p(\{\data_k\}\,|\,\ratepars) &=&
    \int\left[\prod_{k=1}^K p(\data_k\,|\,\entry_k)\right ]
    \,p(\{\entry_k\}\,|\,\ratepars)
    \dd\{\entry_k\}
\end{eqnarray}
where $\{\data_k\}$ is the set of all light curves, \ratepars\ is a vector of
parameters describing the occurence rate function $\rate_\ratepars(\entry)$
and \entry\ is a vector of physical parameters describing a planetary system
(orbital periods, radius ratios, stellar radius, \etc).
In this equation, we have only made one assumption ...

@@

A catalog is a dimensionality reduction of the raw data that with all the
relevant information retained (we hope!).
In the context of \kepler, the catalog reduces the set of downloaded pixel
time series (approximately XXXX data points for the average \kepler\ target)
to probabilistic constraints on a handful of physical parameters---\entry\
from above---like the orbital period and radius.
If we take this set of parameters as \emph{sufficient statistics} of the data
then we can, in theory, compute \eq{crazylike}---up to an unimportant
constant---without ever looking at the raw data again.
This is true because the catalog provides constraints on the posterior
probability of the parameters $\{\entry_k\}$ under some choice of ``interim
prior'' \interim
\begin{eqnarray}\eqlabel{crazypost}
p(\{\entry_k\}\,|\,\{\data_k\},\,\interim) &=&
\frac{p(\{\entry_k\}\,|\,\interim)}{p(\{\data_k\}\,|\,\interim)}\,
\prod_{k=1}^K p(\data_k\,|\,\entry_k) \quad.
\end{eqnarray}
It turns out that we can use this posterior to simplify \eq{crazylike} to a
form that can, in many common cases, be evaluated efficiently.
To find this result, multiply the integrand in \eq{crazylike} by
\begin{eqnarray}
\frac{p(\{\entry_k\}\,|\,\{\data_k\},\,\interim)}
     {p(\{\entry_k\}\,|\,\{\data_k\},\,\interim)}
\end{eqnarray}
and use \eq{crazypost} to find
\begin{eqnarray}
\frac{p(\{\data_k\}\,|\,\ratepars)}{p(\{\data_k\}\,|\,\interim)} &=&
    \int
    \frac{p(\{\entry_k\}\,|\,\ratepars)}{p(\{\entry_k\}\,|\,\interim)}\,
    p(\{\entry_k\}\,|\,\{\data_k\},\,\interim)
    \dd\{\entry_k\} \quad.
\end{eqnarray}
The data only enter this equation through the posterior constraints provided
by the catalog!

@@

In the inference procedure described above, we only discussed an abstract
occurence rate function.
To make it more concrete, we'll model the catalog as a Poisson draw from the
\emph{observable} rate function $\obs{\rate}_\ratepars$.
This leads to the well known result (see \citealt{tabachnik,youdin} for some
of the examples from the exoplanet literature)
\begin{eqnarray}\eqlabel{poisson-like}
p(\{\entry_k\}\,|\,\ratepars) &=&
    \exp\left(-\int \obs{\rate}_\ratepars (\entry) \dd\entry\right) \,
    \prod_{k=1}^K \obs{\rate}_\ratepars (\entry_k)\quad.
\end{eqnarray}
The main thing to note here is that $\obs{\rate}_\ratepars$ is the rate of
exoplanets that you would expect to observe taking into account the geometric
transit probability and any other detection efficiencies.
In practice, it probably makes sense to model the observable rate as
\begin{eqnarray}
\obs{\rate}_\ratepars(\entry) &=&
    \completeness(\entry)\,\rate_\ratepars(\entry)
\end{eqnarray}
where $\completeness(\entry)$ is the detection efficiency at \entry\ and
$\rate_\ratepars(\entry)$ is the object that we want to infer: the true
occurence rate.
Up to this point, we haven't discussed any specific functional form for
$\rate_\ratepars(\entry)$ and all of this derivation is equally applicable
whether we model the occurence rate as (for example) a broken power law or
a histogram.

For the results in this \paper, we will assume that the completeness function
$\completeness(\entry)$ is known empirically but that is not a requirement for
the validity of this method.
Instead, we could use a functional form for it and infer the parameters
(\dfmtodo{cite all the papers that do the ramp, step function, whatever}).

\paragraph{Inverse-detection-efficiency}
It's now interesting to take a brief aside and discuss the connection between
our model and the commonly used inverse-detection-efficiency procedure
(\citealt{howard,dressing,petigura}).
This procedure involves making a weighted histogram of the catalog entries
where the weight for object $\entry_k$ is $1/\completeness(\entry_k)$.
This makes intuitive sense but (to our knowledge) does not have a clear
probabilistic justification or interpretation.

If we model the occurence rate as a histogram with $J$ fixed bin volumes
$\binarea_j$
\begin{eqnarray}
\rate_\ratepars (\entry) &=& \left\{\begin{array}{ll}
\ratepar_1 & \entry \in \binarea_1,\\
\ratepar_2 & \entry \in \binarea_2,\\
\cdots \\
\ratepar_J & \entry \in \binarea_J,\\
0 & \mathrm{otherwise}
\end{array}\right.
\end{eqnarray}
then \eq{poisson-like} becomes
\begin{eqnarray}
\ln p(\{\entry_k\}\,|\,\ratepars) &=&
    \sum_{k=1}^K \sum_{j=1}^J \mathbf{1}[\entry_k \in
        \binarea_j]\,\ln\completeness(\entry_k)\,\ratepar_j
    -\sum_{j=1}^J\ratepar_j\,\int_{\binarea_j} \completeness(\entry)\dd\entry
\end{eqnarray}
where the indicator function $\mathbf{1}[\cdot]$ is one if $\cdot$ is true and
zero otherwise.
If we take the gradient of this function with respect to \ratepars\ and set it
equal to zero, we find the maximum likelihood result
\begin{eqnarray}
\ratepar_j &=& \frac{N_j}{\int_{\binarea_j} \completeness(\entry)\dd\entry}
\end{eqnarray}
where $N_j$ is the number of objects that fall within the bin $j$.
Since it is the maximum likelihood solution, this result is the minimum
variance estimator for the occurence rate in the limit of negligible catalog
uncertainties.
This is \emph{almost} the same as the inverse-detection-efficiency result and
in various limiting cases, they would provide the same result.
\dfmtodo{Finish explaining why this result is better under the specified
assumptions.}

@@

The largest injected radius is $16\,R_\oplus$ but, because of the measurement
uncertainties on the radii, it will be necessary for us to model the
distribution to larger radii.
To do this, we approximate the survey completeness for $\radius>16\,R_\oplus$
as 1.

@@

\acknowledgments
It is a pleasure to thank
    Erik Petigura (Berkeley),
    \etc
for helpful contributions to the ideas and code presented here.
This project was partially supported by the NSF (grant AST-0908357), and NASA
(grant NNX08AJ48G).
This research made use of the NASA \project{Astrophysics Data System}.
